Is essentiality of genes conserved in Enterobacteriaceae?

### 2018-07-06 documents/manuscript.tex
Added an ambiguous region between non-essential and beneficial loss, used eggNOG to study beneficial losses, generated a table containing all KEGG pathways with core essential and ancestrally essential genes
Added not-analysed to DEG analysis

% line numbers
\usepackage[right]{lineno}

@@

% this is required to include graphics
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{pdflscape}
\graphicspath{{../figures/}}

@@

% use for have text wrap around figures
\usepackage{wrapfig}
\usepackage[pscoord]{eso-pic}
\usepackage[fulladjust]{marginnote}
\reversemarginpar
\usepackage{amsmath}

@@

% title goes here:
\begin{flushleft}
{\Large
\textbf\newline{Title}
}
\newline
% authors go here:
\\
%Fatemeh Ashari Ghomi\textsuperscript{1},
%Paul P. Gardner\textsuperscript{1,*},
%Lars Barquist\textsuperscript{2}
%\\
%\bigskip
%\bf{1} University of Canterbury
%\\
%\bf{2} Wurzburg University
%\\
%\bigskip
%* paul.gardner@canterbury.ac.nz

\end{flushleft}

\section*{Abstract}
%Many genes have been identified with advances in sequencing technology and genome annotation methods. However, not all of these genes are of the same importance. We have used transposon mutagenesis to investigate gene essentiality in 14 strains of Enterobacteriaceae. We investigated the potential biases of this approach and found an origin of replication bis, no preferred insertion motif bias, a G-C bias in low G-C genes,  and positional bias within genes. After correcting for these biases, we investigated the changes in the cohorts of essential genes and compared them to their conservation level. Surprisingly, we found that conserved genes are not necessarily essential, and essential genes are not necessarily conserved. However, on average, essential genes are more likely to be conserved.

% now start line numbers
\linenumbers

% the * after section prevents numbering
\section{Introduction}
%The study of cells can give us new insights for understanding the foundation of life. However, the naturally existing cells are very complex which hampers the study of their constituents. Restricting these studies to the genes that are required for the growth of cells can reduce this complexity and enable us to study life in its simplest form \cite{juhas_essence_2011}.

Generating genomic variants that carry genes responsible for essential cellular processes can open up new research directions. Adding genes or metabolic pathways to cells provides new variants with specialised phenotypes. Modified cells have different potential applications in biotechnology \cite{juhas_bacillus_2014}, fuel production \cite{seo_synthetic_2013}, healthcare, and food production \cite{juhas_meeting_2013}. Another important application for studying essential genes is in drug discovery \cite{juhas_essential_2012}. Infectious diseases are among the top major causes of mortality worldwide. Even though antibiotic resistance is growing among bacteria, the antibiotic discovery and development rate is diminishing \cite{fischbach_antibiotics_2009, nathan_antibiotics_2004}. Therefore, it is urgent to find new drugs for infectious diseases. New antibiotics target genes that are essential for the survival of pathogenic bacteria in order to control disease. Some well-known antibiotics that target essential functions are tetracyclines that bind to small ribosomal subunit and interfere with protein translation \cite{brodersen_structural_2000}, penicillins that target peptidoglycan and inhibit the cell wall synthesis \cite{chung_rapid_2009}, and quinolones that target DNA Gyrase \cite{marcusson_interplay_2009}.
%Some candidate antibiotic targets found by investigating essential genes are trans-translation genes \textit{ssrA} and \textit{smpB} from \textit{helicobacter pylori} \cite{thibonnier_trans-translation_2008}, \textit{CcrM} and \textit{MenH} which are bacterial methyltransferases, and \textit{lptD} (previously known as \textit{ostA}) in \textit{Pseudomonas aeruginosa} \cite{srinivas_peptidomimetic_2010} whose essential role is yet under investigation \cite{werneburg_inhibition_2012}.[SOME ANTIBIOTICS THAT TARGET ESSENTIAL GENES ARE...]

In the earliest attempt for the identification of essential genes, Mushegian and Koonin compared the genomes of \textit{Haemophilus influenzae} and \textit{Mycoplasma genitalium}, assuming that the genes that are shared in these two phylogenetically distant bacteria are indispensable and reported 256 genes fulfilling this requirement \cite{mushegian_minimal_1996}. With the advent of sequencing technologies and availability of more genome sequences, the number of core genes in different prokaryotic genomes declined to less than 50 which is not enough for performing all essential functions in a cell \cite{charlebois_computing_2004}. Therefore, the use of experimental methods for the identification of essential genes is vital. Researchers have now studied the essential genes in organisms from all three domains of life \cite{luo_deg_2014} using a number of different methods. Baba et al.\@ \cite{baba_construction_2006} made a library of single gene deletions for \textit{Escherichia coli} K-12. The 303 genes where viable \textit{Escherichia coli} colonies failed to grow are the candidate essential genes. Another group used an antisense RNA knockdown approach to study gene essentiality in \textit{Staphylococcus} aureus \cite{forsyth_genome-wide_2002}. In this method, if the expression of an antisense RNA hinders the growth of the cell, its cognate gene is known as essential. Both of these methods are labour intensive and are dependent on the accuracy of genome annotations. Another widely used procedure is transposon mutagenesis combined with high-throughput sequencing \cite{chao_design_2016, van_opijnen_transposon_2013, barquist_approaches_2013} which includes different approaches namely, Tn-Seq \cite{van_opijnen_tn-seq:_2009}, INSeq \cite{goodman_identifying_2009}, HITS \cite{gawronski_tracking_2009} and TraDIS \cite{langridge_simultaneous_2009}. These procedures differ in the type of transposon, sample preparation methods, and data analysis \cite{van_opijnen_transposon_2013}. Nonetheless, all share the same workflow: pools of single insertion mutants are constructed using transposon mutagenesis. After a growth phase, mutants that are fitter outnumber the less fit ones. Using high-throughput sequencing and tallying transposon junctions gives an indication of whether a genomic region is essential or not. A high number of transposon insertions in a gene indicates that the gene is not essential in its growth medium and conversely, a low number of transposon insertions indicates that a gene is essential in the medium.

There are two groups of essential genes: core essential genes are indispensable for all cells, and accessory essential genes that are required for some organisms. Core essential genes can shed light on the genome structure of the last universal common ancestor and the evolution of living cells \cite{koonin_comparative_2003} and have been used for the synthesis of minimal cells \cite{hutchison_design_2016}. On the other hand, accessory essential genes are helpful in the study of specific lineages. Accessory essential genes may be useful in species-specific antibiotic discovery. Antibiotics that target core essential genes in pathogens may not be ideal, as they may attack homologous genes in their hosts or commensal bacteria.

%Curtis and Brun \cite{curtis_identification_2014} have studied the essentiality changes in cell cycle genes of three alpha-proteobacteria strains: \textit{Caulobacter crescentus}, \textit{Brevundimonas subvibrioides}, and \textit{Agrobacterium tumefaciens} and concluded that although essential genes responsible for cell functions are conserved, there are many essential genes that are specific to each organism.
Freed et al. \cite{freed_combining_2016} have investigated the difference between essential genes in \textit{Shigella flexneri} 2a 2457T and \textit{Escherichia coli} K12 BW25113 and shown that there are no genes that are essential in \textit{Escherichi coli} and not essential in \textit{Shigella flexneri}, while some genes are only essential in \textit{Shigella flexneri}. These include a group of genes involved in cysteine, proline and sugar nucleotide biosynthesis, acetate utilisation, translation elongation, aminoacyl tRNA synthetase, murein DD-endopeptidase, and soxR-reducing complex, many of which are essential in \textit{Shigella flexneri} due to the absence of paralogs or other alternative systems that exist in \textit{Escherichia coli}.
%The difference in essentiality stems from four reasons: the disruption of many of the genes in \textit{Shigella flexneri} 2a 2457T impede the growth, but these genes are not actually essential; the conditions were different in the two experiments done on \textit{Escherichia coli} K12 BW25113 and \textit{Shigella flexneri} 2a 2457T; many of the genes were not actually essential for growth, but essential for transposon insertion and antibiotic resistance; and

%PPG: Reduce, summarise main findings of Canals & Barquist. e.g. Most essential genes are also essential in related species, yet there are some serovar-specific essential genes [GIVE SOME NUMBERS]... 
Canals et al.\@ \cite{canals_high-throughput_2012} have compared the essentiality of genes in \textit{Salmonella} typhimurium ATCC 14028, two isolates of \textit{Salmonella typhi} Ty2 (varying in \textit{htrA}, \textit{aroC} and \textit{aroD} genes), and \textit{Escherichia coli} K12 BW25113 and found that these cells share 268 essential genes. Nine genes are essential in \textit{Escherichia coli} and not essential in \textit{Salmonella}s, three of which can tolerate insertions only in some parts and one has distant paralogs in \textit{Salmonella}s that is missing in \textit{Escherichia coli}. For other genes, there are a few insertions in \textit{Salmonella}s, but the number is not small enough to call those genes essential. Moreover, 159 genes were almost essential in all \textit{Salmonella}s but not in \textit{Escherichia coli}. These include genes involved in replication and genes related to ribosome and its accessory proteins. The authors also found 26 genes that are under greater selection in \textit{Salmonella} Typhimurium compared to \textit{S. typhi} and 10 genes vice versa. Barquist et al. \cite{barquist_comparison_2013} have used transposon-directed insertion-site sequencing to compare the essentiality of genes in \textit{Salmonella} Typhi Ty2, \textit{Salmonella} Typhimurium SL1344, and \textit{Escherichia coli} K12 BW25113. 
%THIS IS IMPORTANT:
These genomes share 228 essential genes which are mostly involved in cell division, transcription, translation, and fatty acid and peptidoglycan biosynthesis. Additionally, many of the serovar-specific essential genes in \textit{Salmonella}s are phage repressors. Another key difference between these two serovars is that a set of genes that are putatively involved in cell wall biosynthesis are essential in \textit{Salmonella} Typhimurium and not essential in \textit{salmonella} Typhi which gives an indication of the adaptation of these two \textit{Salmonella} serovars to their niches.

Although there are some studies on differentiation of essentiality in different organisms, these studies usually include a few genomes. We are going to extend these studies to a larger scale in Enterobacteriaceae family. Our aim is to study the essentiality of genes in in 16 different organisms. These include \textit{Enterobacter cloacae} NCTC 9394, \textit{Klebsiella pneumoniae} Ecl8, \textit{Klebsiella pneumoniae} RH201207, \textit{Citrobacter rodentium} ICC168, \textit{Salmonella} Typhimurium SL1344, \textit{Salmonella} Typhimurium SL3261, \textit{Salmonella} Typhimurium D23580, \textit{Salmonella} Typhimurium A130, \textit{Salmonella} Enteritidis P125109, \textit{Salmonella} Typhi Ty2, \textit{Escherichia coli} ST131 EC958, \textit{Escherichia coli} UPEC ST131, \textit{Escherichia coli} ETEC CS17, \textit{Escherichia coli} ETEC H10407, \textit{Escherichia coli} BW25113, and \textit{Escherichia coli} K-12 MG1655.

%WHAT FATEMEH DID....

%Enterobacteriaceae is a well characterised family of Gram-negative bacteria with a variety of host ranges and pathogenecity \cite{brenner_bergeys_2006}. In addition, we added the essential genes of \textit{Escherichia coli} K-12 MG1655 from EcoGene database \cite{zhou_ecogene_2013} to our study. In EcoGene the essentiality of genes suggested as essential by Baba et al.\@ \cite{baba_construction_2006} has been further studied and only 299 out of 303 genes are marked as essential. We first performed a detailed study of biases than can influence the inference of essentiality. Then, we normalised our data for the biases and investigated the essentiality of genes in three classes of genes: genus specific (genes that are present only in one genus), single copy genes (genes with about one instance per genome in all of the genomes that we are studying), and multi-copy genes(genes that are copied multiple times per genome). We have also investigated how essentiality changes in the phylogenetic tree for these organisms.

%\begin{figure}
%\includegraphics[scale=0.2]{phylosift-aa-raxmlbootstrap.pdf}
%\caption{The species tree containing the 16 genomes used in this study. We have generated the tree by running RAxML \cite{stamatakis_raxml_2014} on Phylosift \cite{darling_phylosift:_2014} amino acid markers.}
%\label{fig:species-tree}
%\end{figure}

\section{Results and Discussion}
\subsection{Incorporating multiple measures of gene essentiality improves prediction}
In this section we will compare a number of methods of quantifying the essentiality of genes based on transposon insertion data and introduce our own method.
\subsubsection{Comparison of old methods}
A number of methods have been used for evaluating the essentiality of genes using transposon insertion data. Freed et al. \cite{freed_combining_2016} have compared eleven features that can quantify the essentiality of genes. These include: the number of insertion sites within genes; the mean distance between insertion sites, the median distance between insertion sites, the number of base pairs before the first insertion in the $5^\prime$ end, the ratio between the number of base pairs before the first insertion in the $5^\prime$ end and gene length, the number of base pairs before the second insertion in the $5^\prime$ end, the ratio between the number of base pairs before the second insertion in the $5^\prime$ end and gene length, the length of the longest uninterrupted region in the gene, the ratio between the length of the longest uninterrupted region in the gene and gene length, the length of the longest region in the gene containing at most one insertion divided by gene length, and the number of insertion sites divided by gene length. Among these, the average distance between insertion sites and the length of the longest uninterrupted region were the best predictors.

Barquist et al. \cite{barquist_tradis_2016} have used insertion indices which are calculated by dividing the number of insertion sites by gene length for defining the essentiality of genes. Plotting all insertion indices for all genes in a genome gives a bimodal plot, each mode representing a group of genes (essential or non-essential). They have fitted two gamma distributions to these modes and calculated the log odds value for each gene to test which mode it belongs to.

Turner et al. \cite{turner_essential_2015} have randomly sampled insertions in a genome and using DESeq package \cite{anders_differential_2010} calculated log fold changes and P-values for the actual number of insertions compared to the expected number of insertions obtained from the sampling method.

We have compared the predictive power of the average distance between insertion sites in a gene, the length of the longest uninterrupted region, insertion index, log odds value obtained from insertion index value, and P-value and log fold change calculated using DESeq package after sampling. To evaluate the accuracy of these methods, we have compared the predicted essential genes to the essential genes in \textit{Escherichia coli} K-12 MG1655 in EcoGene database \cite{zhou_ecogene_2013}. The results are depicted in Figure~\ref{fig:benchmark}. Among all methods, those that try to fit a distribution to data and predict essentiality based on that, namely log odds value from insertion indices and P-value from DESeq after sampling, are the least accurate methods. Sampling and then calculating log fold changes using DESeq package is the most accurate predictor and the other three methods are also good predictors despite their simplicity.
\subsubsection{PCA based method}\label{sec:essentiality-call}
We selected four of the predictors in previous section: insertion index, log fold change using sampling and DESeq, the average distance between insertion sites in a gene, and the length of the longest uninterrupted region. These predictors were selected as they were more accurate and less dependent on each other. Afterwards, we ran principal component analysis on these predictors using \emph{prcomp} function in R and selected the first principal component. As Figure~\ref{fig:benchmark} shows, PCA results gave us the most powerful tool for quantifying the essentiality of genes compared to other methods. We have plotted the results for one sample genome and the results for other genomes are shown in {\color{red}SUPPLEMENTARY}. Figure~\ref{fig:pca} shows the distribution of PCA results after zero mean unit variance normalisation. We call this value NPEQ (Normalised Pca based Essentiality Quantifier) in this paper. The genes whose NPEQ values are less than -1.644854 after normalisation are beneficial losses and the genes with NPEQ greater than 1.644854 are essential. The rest are non-essential genes. The cutoff 1.644854 has been selected because if we assume that NPEQ distribution is normal, this cutoff shows the P-value 0.95. Moreover, by trying to define a cutoff that maximises Matthews correlation coefficient for each genome using NPEQ, we get 1.650449 as the average cutoff which is very close to this value.

\begin{figure}
\captionsetup[subfigure]{justification=centering}
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[page=1, scale=0.4]{essential-call-comparison-BW25113.pdf}
  \caption{}
  \label{fig:benchmark}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[page=10, scale=0.4]{essential-call-comparison-BW25113.pdf}
  \caption{}
  \label{fig:pca}
\end{subfigure}
\caption{(a) The accuracy of 7 different prediction methods for quantifying the essentiality of genes. The higher the area under the curve, the more accurate the method is. (b) The distribution for our proposed method (NPEQ)}
\label{fig:essentiality-call}
\end{figure}

%\begin{figure*}
%\centering
%\begin{tabular}{c c c}
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-BN373.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-BW25113.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-CS17.pdf}\\
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-EC958.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ENC.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ERS227112.pdf}\\
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ETEC.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-NCTC13441.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ROD.pdf}\\
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-SEN.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-SL1344.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-SL3261.pdf}\\
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-STM.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-STMMW.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-t.pdf}\\
%\end{tabular}
%\caption{The plots show the accuracy of 7 different prediction methods for quantifying the essentiality of genes. The higher the area under the curve, the more accurate the method is.}
%\label{fig:benchmark}
%\end{figure*}

>>>>

% line numbers
\usepackage[right]{lineno}
\linenumbers

@@

% this is required to include graphics
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{pdflscape}
\graphicspath{{../figures/panels/}}

@@

% use for have text wrap around figures
\usepackage{wrapfig}
\usepackage[pscoord]{eso-pic}
\usepackage[fulladjust]{marginnote}
\reversemarginpar
\usepackage{amsmath}
\usepackage{pdfpages}

@@

% title goes here:
\begin{flushleft}
{\Large
\textbf{Investigating essential genes in Enterobacteriaceae using
transposon insertion sequencing}
}
\newline
% authors go here:
\\
%Fatemeh Ashari Ghomi\textsuperscript{1},
%Paul P. Gardner\textsuperscript{1,*},
%Lars Barquist\textsuperscript{2}
%\\
%\bigskip
%\bf{1} University of Canterbury
%\\
%\bf{2} Wurzburg University
%\\
%\bigskip
%* paul.gardner@canterbury.ac.nz

\end{flushleft}% Correct contributions, italicise bacteria names, italicise gene names, abbreviate species names
% convert MG1655(keio) to BW25113    http://mbio.asm.org/content/6/3/e00306-15.full#sec-11
% EC958 -> CFT073    http://journals.plos.org/plospathogens/article?id=10.1371/journal.ppat.1003788#ppat.1003788.s003
\section{Abstract}
The study of genes that are essential for the survival of a prokaryotic organism can help with the design of new antibiotics and designing minimal viable genomes that can be ``programmed'' with new genes. Transposon insertion is a method for defining essential genes. We have used this method for different bacteria in the Enterobacteriaceae family. Here, we show that on average a density of one insertion site every 25 nucleotides can result in accurate prediction of essential genes and introduce a new computational method for defining essential genes using transposon insertion data. In addition, we studied possible sources of bias in this experiment. These analyses show that there is a decline in the number of transposon insertions as we get further from the origin of replication. Moreover, insertions in the $5^\prime$ and $3^\prime$ ends of genes do not follow the same pattern as intermediate region. We compared the essential genes in Enterobacteriaceae with other bacteria and found that genes that are essential in all bacteria are involved in replication, transcription, translation, and cell division. However, many genes involved in these processes are not necessarily essential in all bacteria due to the existence of alternatives that compensate for them. In addition, we show that the essentiality of genes and their conservation are related: essential genes are more likely to be conserved and conserved genes are more likely to be essential.
%\section{Contributions}
%Lars Barquist and Paul Gardner have supervised the project and provided feedback on the manuscript. I have performed the analyses and written the manuscript. The transposon insertion sequencing data has been provided by Amy Cain, Christine Boinett, Gemma Langridge, Julian Parkhill, and Nicholas Thomson at Wellcome Sanger Institute, Moataz Abd El Ghany at King Abdullah University of Science and Technology, and Lars Barquist at Helmholtz Institute for RNA-based Infection Research.
\section{Introduction}
Sequencing a new genome and annotating the genomic regions is now quick and easy thanks to advances in sequencing and genome annotation methods, however the functional annotation of genomes remains challenging. One aspect of functional annotation is the study of genes that are needed for the viability of an organism i.e. essential genes. Detecting essential genes  is important for furthering our understanding of fundamental biology. One of the applications of identifying essential genes is drug discovery \cite{juhas_essential_2012}. New antibiotics, target genes that are essential for the survival of pathogenic bacteria in order to control disease. Moreover, essential genes can serve as the basis for generating minimal genomes \cite{glass_essential_2006, hutchison_global_1999}. Minimal genomes are appealing for two reasons. First, to understand how small can minimal life be and second, for improving biotechnological processes \cite{martinez-garcia_quest_2016}.

In the earliest attempt for the identification of essential genes, Mushegian and Koonin compared the genomes of \textit{Haemophilus influenzae} and \textit{Mycoplasma genitalium}, assuming that the genes that are shared in these two phylogenetically distant bacteria are indispensable and reported 256 genes fulfilling this requirement \cite{mushegian_minimal_1996}. With the advent of sequencing technologies and availability of more genome sequences, the number of core genes in different prokaryotic genomes has declined to less than 100 \cite{brown_universal_2001, koonin_comparative_2003, harris_genetic_2003}. This number does not seem to be enough for performing all essential functions in a cell since many of the essential functions can be performed by non-orthologous genes. Therefore, computational methods are not capable of defining essential genes and the use of experimental methods for the identification of essential genes is important. By combining computational and experimental studies, Gil et al. introduced a set of 206 genes that can theoretically be used to design a genetically minimal bacterium \cite{gil_determination_2004}. They have divided the genes responsible for essential functions into 5 groups: 1) information storage and processing which can be related to DNA or RNA metabolism, 2) protein processing, folding, and secretion, 3) cell structure and cellular processes, 4) energetic and intermediary metabolism, and 5) poorly characterised genes. However, in practice the number of genes that appear to be required for the smallest viable synthetic bacterium is 437 \cite{hutchison_design_2016}.

Essential genes have been studied in organisms from all three domains of life \cite{luo_deg_2014} using a number of different methods. Baba et al. \cite{baba_construction_2006} made a library of single gene deletion mutants for \textit{E. coli} K-12 BW25113, called Keio collection, and identified 303 genes that hindered the growth of viable \textit{E. coli} colonies as candidate essential genes.  Antisense RNA knockdowns have also been used to study gene essentiality in \textit{Staphylococcus aureus} \cite{forsyth_genome-wide_2002}. With this approach, if the expression of an antisense RNA hampers cell growth, its cognate gene is thought to be essential. Both methods are labour intensive and depend on the accuracy of genome annotation. A less targeted procedure for identifying essential genes is transposon mutagenesis combined with high-throughput sequencing \cite{barquist_approaches_2013, van_opijnen_transposon_2013, chao_design_2016} which is used in several different approaches, namely Tn-Seq \cite{van_opijnen_tn-seq:_2009}, INSeq \cite{goodman_identifying_2009}, HITS \cite{gawronski_tracking_2009} and TraDIS \cite{langridge_simultaneous_2009}. These methods differ in the type of transposon, sample preparation methods, and data analysis \cite{van_opijnen_transposon_2013}. Nonetheless, all share a similar workflow: pools of single insertion mutants are constructed using transposon mutagenesis and screened using a selective media for the inserted transposons. After a recovery phase and library construction, high-throughput sequencing is used to identify transposon insertion junctions, which are then tallied to indicate whether a genomic region is essential or not. A high number of transposon insertions in a gene indicates that the gene is not essential in its growth medium, and conversely a low number of transposon insertions indicates that a gene is essential in the medium.

The study of core essential genes i.e. genes that are essential in all bacteria can shed light on the genome structure of the last universal common ancestor and the evolution of living cells \cite{koonin_comparative_2003}, and provide insight into the requirements for the synthesis of a universal minimal cell. Knowledge of accessory essential genes, which are only essential in specific lineages, may help identify species-specific antibiotic targets which would not adversely affect the host or commensal bacteria. A number of studies have used transposon insertion experiments to compare cohorts of essential genes and showed that closely related organisms can have specific essential genes. Freed et al. \cite{freed_combining_2016} have investigated the difference between essential genes in \textit{Shigella flexneri} and \textit{E. coli} K-12 BW25113 and shown that there are no genes that are essential in \textit{E. coli} and not essential in \textit{S. flexneri}, while some genes are only essential in \textit{S. flexneri}. These include a group of genes involved in cysteine, proline and sugar nucleotide biosynthesis, acetate utilisation, translation elongation, aminoacyl tRNA synthetase, murein DD-endopeptidase, and the soxR-reducing complex. Many of these are essential in \textit{S. flexneri} due to the absence of paralogs or other alternative pathways that exist in \textit{E. coli}.

In another study, Canals et al. \cite{canals_high-throughput_2012} have compared the essentiality of genes in \textit{Salmonella} Typhimurium, two isolates of \textit{S.} Typhi Ty2 (varying in \textit{htrA}, \textit{aroC} and \textit{aroD} genes), and \textit{E. coli} K-12 BW25113. Overall, 268 essential genes were found to be shared between these strains. Nine genes are essential in \textit{E. coli} and not essential in \textit{Salmonella}. These genes are mostly near essential in \textit{Salmonella} as well. %, three of which can tolerate insertions only in some parts and one has distant paralogs in Salmonellas that is missing in \textit{E. coli}. For other genes, there are a few insertions in Salmonellas, but the number is not small enough to call those genes essential.
Moreover, 159 genes were essential in \textit{Salmonella} but not in \textit{E. coli}. These include genes involved in replication and genes related to ribosome and its accessory proteins. The authors also found 26 genes that are under greater selection in \textit{S.} Typhimurium compared to \textit{S.} Typhi and 10 genes vice versa. Barquist et al. \cite{barquist_comparison_2013} have used transposon-directed insertion-site sequencing to compare the essentiality of genes in \textit{S.} Typhi, \textit{S.} Typhimurium, and \textit{E. coli} K-12 BW25113. These genomes share 228 essential genes which are mostly involved in cell division, transcription, translation, and fatty acid and peptidoglycan biosynthesis. Additionally, many of the serovar-specific essential genes in \textit{Salmonella} are phage repressors. These proteins keep the phages in lysogenic phase and stops them from entering the lytic cycle that destructs the cell \cite{echols_establishment_1971}. Another key difference in the two \textit{Salmonella} serovars is the differences in the essentiality of genes involved in iron utilisation and cell surface structure that gives an indication of the adaptation of these two \textit{Salmonella} serovars to their niches. %that a set of genes that are putatively involved in cell wall biosynthesis are essential in \textit{S.} Typhimurium and not essential in \textit{S.} Typhi which gives an indication of the adaptation of these two \textit{Salmonella} serovars to their niches.

Although there are some studies on differentiation of essentiality in different organisms, their scope is limited to a handful of species/genomes. We aimed to extend these studies to a larger scale by investigating the essentiality of genes in 14 different organisms in the Enterobacteriaceae family of bacteria using TraDIS. These bacteria include \textit{Enterobacter cloacae} NCTC 9394, \textit{Klebsiella pneumoniae} Ecl8, \textit{K. pneumoniae} RH201207, \textit{Citrobacter rodentium} ICC168, \textit{S.} Typhimurium SL1344, \textit{S.} Typhimurium SL3261, \textit{S.} Typhimurium D23580, \textit{S.} Typhimurium A130, \textit{S.} Enteritidis P125109, \textit{S.} Typhi Ty2, \textit{E. coli} UPEC CFT073, \textit{E. coli} UPEC ST131, and \textit{E. coli} K-12 BW25113 which is the parent of \textit{E. coli} strain used in Keio collection \cite{baba_construction_2006}. We have also used the essentiality data of \textit{E. coli} K-12 BW25113 from EcoGene database \cite{zhou_ecogene_2013} which improves the essentiality data in Keio collection \cite{baba_construction_2006}. We first introduce a computational method for calling essential genes, compare it with the existing methods and investigate the biases that may affect our results. In addition, we find the best insertion density for transposon insertion experiments in bacteria. We then compare the list of core and accessory essential in different lineages and investigate the implications. %genes to core essential genes in all bacteria and core genes in Gammaproteobacteria endosymbionts to find which genes are essential in which lineages. 
\section{Results and discussion}
In this section we will compare different methods for evaluating gene essentiality and describe our approach. Then, we will discuss the ideal insertion rate for transposon insertion experiments. Furthermore, we will review different sources of bias in transposon insertion experiments and study the essentiality of genes in different bacterial lineages.
\subsection{Benchmarking different essentiality evaluators}
The number of transposon insertions detected in a gene is negatively correlated with the essentiality level of that gene. The more insertions exist in a gene, the less essential that gene is. A number of methods have been used for evaluating the essentiality of genes using transposon insertion data. Barquist et al. \cite{barquist_tradis_2016} use insertion indices which are calculated by dividing the number of insertion sites in a gene by gene length. Plotting the insertion indices for all genes in a genome gives a bimodal plot (Figure~\ref{fig:fig1}a), each mode representing a group of genes (essential or non-essential). They have fitted two gamma distributions to these modes and calculated the log odds value to select a cut-off on the insertion index that distinguishes between essential and non-essential genes.

Freed et al. \cite{freed_combining_2016} compared eleven features that can quantify the essentiality of genes, among which the average distance between insertion sites and the largest uninterrupted fraction were the best predictors. Turner et al. \cite{turner_essential_2015} used Monte Carlo method to randomly simulate insertions in a genome and using the DESeq package \cite{anders_differential_2010} calculated log fold changes for the actual number of insertions compared to the expected number of insertions obtained from the randomisation method, and clustered the log fold changes using Mclust package in R \cite{scrucca_mclust_2016}.

We compared the predictive power of the average distance between insertion sites in a gene, the largest uninterrupted fraction, insertion index, and Monte Carlo method combined with DESeq. To evaluate the accuracy of these methods, we compared essential genes predicted with each method to the essential genes in \textit{E. coli} K-12 BW25113 in the EcoGene database \cite{zhou_ecogene_2013}. In this database, the essential genes from Keio collection \cite{baba_construction_2006} are further studied and some are omitted from the list (e.g. a non-essential signal peptide needed for the translation of an essential gene, or pseudogenes) while some other genes are added (mostly due to the existence of other copies that compensate for them when deleted \cite{yamamoto_update_2009}). The results of our comparison are depicted in Figure~\ref{fig:fig1}b and Supplementary Figure 1. Most of these methods have very close accuracies, but on average, insertion index is the most accurate predictor as it has the highest average Area Under the Curve (AUC) over all 13 datasets (Supplementary Figure 2). We also combined all these methods using Principal Component Analysis (PCA). We ran the prcomp function in R and selected the first principal component for our analysis. The average AUC is 0.9608 for PCA. As the random sampling method is too slow, our PCA method runs very slowly with only a small improvement to the results. Therefore, we tried PCA excluding random sampling which resulted in an average AUC higher than insertion index (Supplementary Figure 2). The difference between average AUCs of PCA excluding random sampling and insertion index is small and insertion index is a straightforward and more intuitive measure. Therefore, we used insertion index for identifying essential genes. As our dataset is composed of the results of different transposon insertion experiments on different genomes, and the number of transposon insertion sites and genome length varies in these experiments, we have divided the insertion indices in each genome by $n_o/l_o$ where $n_o$ is the number of insertion sites in genome and $l_o$ is the length of the genome. The distribution of the insertion indices for \textit{E. coli} BW25113 is plotted in Figure~\ref{fig:fig1}a.

Another challenge is to define a threshold for separating essential genes from non-essential ones. One method, is using gamma fits to find this threshold \cite{barquist_tradis_2016}. However, with this method, when gamma fits are not consistent with the sample distributions, the log odds values obtained from them are not good predictors of essentiality. In order to make our predictions independent of the goodness of fits, we have used DBSCAN clustering \cite{ester_density-based_1996}. The details of this method are provided in Methods~\ref{meth:essentiality}. This method clusters genes whose insertion indices are closely packed together and results in four groups of genes: essential genes, non-essential genes, beneficial losses, and ambiguous genes which are between different groups. Supplementary Figure 3 compares the results of essential gene prediction using gamma fits and DBSCAN. In this figure, we have assumed that essential genes in \textit{E.coli} K-12 BW25113 which are proposed in EcoGene are also essential in other genomes and calculated Matthews Correlation Coefficient for different cut-offs on insertion index. The cut-off found by DBSCAN outperforms gamma fits in predicting essential genes.

\begin{figure}
\includegraphics[scale=0.45]{fig1.pdf}
\caption{(a) The bimodal distribution of insertion indices shows different classes of essentiality obtained using DBSCAN. These include essential genes which have the lowest insertion indices, non-essential genes with intermediate insertion indices, beneficial losses with high insertion indices, and ambiguous genes which are located between these groups (b) ROC curves showing the accuracy of 6 methods for predicting essential genes. True positives are genes that are predicted as essential and whose orthologs are essential in EcoGene database of essential genes for \textit{E. coli} K-12 BW25113. False positives are genes that are predicted as essential, but are classed as non-essential in EcoGene study.}
\label{fig:fig1}
\end{figure}

\subsection{What is an ideal insertion density?}
If the number of transposon insertions is too low compared to genome length, the insertion index distribution is not bimodal and therefore the prediction of essential genes is not accurate. However, higher numbers of insertions are labour intensive and cost more due to larger mutant pools which need to be sequenced. Insertion density is defined as the ratio between the number of insertion sites and genome length. In this section, we explore the minimum insertion density which does not affect the accuracy of the experiment.

We provided 4 independent \textit{S.} Typhimurium SL1344 libraries with 50K, 100K, 250K, 500K, and 1000K Colony-Forming Unit (CFU). On average, these libraries have 4075282, 4160203, 4526237, 5711423, and 13813802 reads respectively which results in one insertion site in every 57.07, 52.94, 16.49, 14.17 and 8.86 nucleotides, assuming a uniform distribution of insertions. In order to compare different ranges of insertion densities, we simulated different libraries with one insertion site in 8.86 to 243.90 nucleotides. For each simulated library, we used the closest real library with more insertion sites than our simulated library and randomly omitted insertion sites until we reached the desired density.

The false positive rate declines with the increase of insertion density as illustrated in Figure~\ref{fig:fig2}a and at 0.04 density it becomes almost constant. Figure ~\ref{fig:fig2}b shows that the true positive rate goes down as the insertion density rises to 0.01. This happens due to the large number of predicted essential genes which also includes many actual essential genes. But, as the density increases the number of predicted essential genes decreases which results in a decline in true positive rate. After 0.01, the true positive rate rises with increasing insertion density and it gets almost constant at about 0.04 insertion density. As a result of the fluctuations of the true positive rate and the false positive rate, Matthew's Correlation Coefficient goes up with higher insertion densities until it reaches 0.04. Then, it changes unnoticeably with more insertions (Supplementary Figure 4a). Finally, the percentage of genes with at least 1 insertion rises with higher insertion densities, but the curve becomes almost flat after 0.04 (Supplementary Figure 4b). Overall, these figures show that an insertion density about 0.04 can give us a good trade-off between cost and accuracy.
